\Large The psychological arrow of time drives temporal asymmetries in inferring unobserved past and future events

### 2023-09-12 paper/old.tex
working on addressing reviewer comments [WIP]

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{methods}
  \caption{\textbf{Task overview.} Participants watched segments of two storylines from the television series \textit{Why Women Kill}.  They made free-form text responses to either retrodict what had happened in the previous segment, predict what would happen in the next segment, or recall what happened in the just-watched segment. Across four counterbalanced sequences, we systematically varied whether participants watched the segments in forward or reverse chronological order, whether (or not) responses were cued using the main characters in the target segment, and which other segments participants had watched prior to making a response.  For each segment, we collected several retrodiction, prediction, and/or recall responses across different experimental conditions.  Experiment time is denoted along the
  vertical axis, storyline segments are indicated along the horizontal axis, and the colors denote experimental tasks (conditions).}
  \label{fig:method}
\end{figure}

@@

If characters' direct references cannot fully account for the temporal asymmetry in retrodicting the unobserved past versus predicting the unobserved future, what other factors might explain this phenomenon?  The results above indicate that characters’ references to specific unobserved events in the past or future boost participants’ estimates of these events.  But might characters' references have other effects on participants' responses \textit{beyond} the referenced events?  For example, real-world experiences and events in realistic narratives are often characterized by temporal autocorrelations (i.e., what is ``happening now'' will likely relate to what happens ``a moment from now,'' and so on). Real-world experiences and realistic narratives are also often structured into ``schemas'' whereby experiences unfold according to a predictable pattern or formula that characterizes a particular situation, such as going to a restaurant or catching a flight at the airport~\citep{BaldEtal18}.  If there are associations or temporal dependencies between temporally nearby events in the television show participants watched, participants might be able to pick up on these patterns in forming their responses.  This would be reflected in an inference ``boost'' for events that were \textit{nearby in time} to events that characters referred to in their conversations, in addition to the referenced events themselves (Fig.~\ref{fig:result4}A).

Because characters tended to refer to past events more often than future events, the proportions of unreferenced events that were adjacent to referenced events should show a similar temporal asymmetry in favor of the past.  We tested this intuition by computing the proportions of unreferenced events in the stimulus that were temporally adjacent to past or future events referenced by the characters during a given segment.  Here we defined \textit{temporally adjacent} as any event within an absolute lag of one relative to a referenced onscreen event, or within an absolute lag of 0.5 to a referenced offscreen event.  We also defined \textit{remaining} events as unreferenced events that were not temporally adjacent to any referenced events.  As shown in Figure~\ref{fig:result4}B, we observed higher proportions of unreferenced past than future events that were temporally adjacent to referenced events.  Further, these reference-adjacent events had higher hit rates than remaining events after controlling for absolute lag (uncued retrodiction: OR = 7.15, $Z = 2.40$, $p = 0.02$, CI: 1.44 to 35.58; uncued prediction: OR = 3.11, $Z = 2.30$, $p = 0.02$, CI: 1.18 to 8.21; Fig.~\ref{fig:result4}E).  To estimate the contributions of reference adjacency on hit rates, we computed the difference in hit rates between unreferenced events (which comprised both reference-adjacent and remaining events) and remaining events, as a function of lag.  These differences exhibited a temporal asymmetry in favor of retrodiction.  This suggests that reference-adjacent events also contribute to participants' retrodiction advantage.  Remaining events did \textit{not} exhibit a reliable temporal asymmetry (OR = 0.75, $Z = 0.33$, $p = 0.74$, CI: 0.14 to 4.08; Fig.~\ref{fig:result4}D), suggesting that, after accounting for temporal adjacency, character's references to past and future events can explain participants' retrodiction advantage.

>>>>

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{methods}
  \caption{\textbf{Task overview.} Participants watched segments of two storylines from the television series \textit{Why Women Kill}.  They made free-form text responses to either retrodict what had happened in the previous segment, predict what would happen in the next segment, or recall what happened in the just-watched segment. Across four counterbalanced sequences, we systematically varied whether participants watched the segments in forward or reverse chronological order, whether (or not) responses were cued using the main characters in the target segment, and which other segments participants had watched prior to making a response.  For each segment, we collected several retrodiction, prediction, and/or recall responses across different experimental conditions.}
  \label{fig:method}
\end{figure}

@@

If characters' direct references cannot fully account for the temporal asymmetry in retrodicting the unobserved past versus predicting the unobserved future, what other factors might explain this phenomenon?  The results above indicate that characters’ references to specific unobserved events in the past or future boost participants’ estimates of these events.  If there are associations and dependencies between temporally adjacent events, might characters’ references to specific events also boost participants’ estimates of other events that were temporally \textit{adjacent} to the referenced events (Fig.~\ref{fig:result4}A)?  Because characters tended to refer to past events more often than future events, the proportions of unreferenced events that were adjacent to referenced events should show a similar temporal asymmetry in favor of the past.  We tested this intuition by computing the proportions of unreferenced events in the stimulus that were temporally adjacent to past or future events referenced by the characters during a given segment.  Here we defined \textit{temporally adjacent} as any event within an absolute lag of one relative to a referenced onscreen event, or within an absolute lag of 0.5 to a referenced offscreen event.  We also defined \textit{remaining} events as unreferenced events that were not temporally adjacent to any referenced events.  As shown in Figure~\ref{fig:result4}B, we observed higher proportions of unreferenced past than future events that were temporally adjacent to referenced events.  Further, these reference-adjacent events had higher hit rates than remaining events after controlling for absolute lag (uncued retrodiction: OR = 7.15, $Z = 2.40$, $p = 0.02$, CI: 1.44 to 35.58; uncued prediction: OR = 3.11, $Z = 2.30$, $p = 0.02$, CI: 1.18 to 8.21; Fig.~\ref{fig:result4}E).  To estimate the contributions of reference adjacency on hit rates, we computed the difference in hit rates between unreferenced events (which comprised both reference-adjacent and remaining events) and remaining events, as a function of lag.  These differences exhibited a temporal asymmetry in favor of retrodiction.  This suggests that reference-adjacent events also contribute to participants' retrodiction advantage.  Remaining events did \textit{not} exhibit a reliable temporal asymmetry (OR = 0.75, $Z = 0.33$, $p = 0.74$, CI: 0.14 to 4.08; Fig.~\ref{fig:result4}D), suggesting that, after accounting for temporal adjacency, character's references to past and future events can explain participants' retrodiction advantage.
